An interdisciplinary examination of stress and injury occurrence in athletes

### 2020-07-20 full_paper_v2.tex
pre results edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.
%                                              %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.4 Generated 2018/06/15 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

\documentclass[utf8]{frontiersHLTH}

%\setcitestyle{square} % for Physics and Applied Mathematics and Statistics articles
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}

\linenumbers

% BELOW TAKEN FROM rticles plos template
%
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% hyperref package, useful for hyperlinks
\usepackage{hyperref}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% Sweave(-like)
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}
\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Below is from frontiers
%
\bibliographystyle{frontiersinSCNS}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace}
%\doublespacing

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

%% END MACROS SECTION

\def\Authors{
  Dr Harry Fisher\,\textsuperscript{1*},
  Dr Marianne Gittoes\,\textsuperscript{1},
  Professor Lynne Evans\,\textsuperscript{1},
  Miss Leah Bitchell\,\textsuperscript{1},
  Dr Richard Mullen\,\textsuperscript{2},
  Dr Marco Scutari\,\textsuperscript{3}}

\def\Address{

  \textsuperscript{1} School of Sport and Health Science, Cardiff Metropolitan University,  Cardiff,   United Kingdom

  \textsuperscript{2} Division of Sport, Health \& Exercise Sciences, Brunel University,  London,   United Kingdom

  \textsuperscript{3} Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA),  Manno,   Switzerland
  }

  \def\corrAuthor{Dr Harry Fisher}\def\corrAddress{Cardiff Metropolitan University\\Cyncoed Road\\, CF23 6XD United Kingdom}\def\corrEmail{\href{mailto:harryfisher21@gmail.com}{\nolinkurl{harryfisher21@gmail.com}}}
  \def\firstAuthorLast{Fisher {et~al.}}

\begin{document}
\onecolumn
\firstpage{1}

\title[Short Title]{An interdisciplinary examination of stress and injury occurrence in
athletes}
\author[\firstAuthorLast]{\Authors}
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

@@

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

Over the last four decades sport related injuries have received
increased research attention (Ivarsson et al., 2017). This attention is
unsurprising given the high incidence (Rosa et al., 2014; Sheu et al.,
2016), and undesirable physical and psychological effects of sports
injuries (Leddy et al., 1994; Brewer, 2012). To mitigate against both
the increasing incidence and undesirable consequences of injury,
research has identified several psychological (Slimani et al., 2018),
anatomical (Murphy et al., 2003), biomechanical (Neely, 1998; Hughes,
2014) and environmental (Meeuwisse et al., 2007) factors associated with
sports injury occurrence. Indeed, several models of injury causation
have been proposed that highlight the multifactorial nature of injury
occurrence (Kumar, 2001; Meeuwisse et al., 2007; Wiese-Bjornstal, 2009),
of which one of the most widely cited was developed by Williams and
Anderson (Fig \ref{fig:fig1}; Andersen and Williams, 1988; Williams and
Andersen, 1998).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{frontiersHLTH}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={An interdisciplinary examination of stress and injury occurrence in athletes},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{An interdisciplinary examination of stress and injury occurrence
in athletes}
\author{true \and true \and true \and true \and true \and true}
\date{}

\begin{document}
\maketitle

@@

\section{Introduction}

Over the last four decades sport-related injuries have received
increased research attention (Ivarsson et al., 2017) in response to the
high incidence (Rosa et al., 2014; Sheu et al., 2016) and associated
undesirable physical and psychological effects (Leddy et al., 1994;
Brewer, 2012). Multiple psychological (Slimani et al., 2018), anatomical
(Murphy et al., 2003), biomechanical (Neely, 1998; Hughes, 2014) and
environmental (Meeuwisse et al., 2007) factors have been associated with
sports injury occurrence and several models of injury causation have
been proposed that highlight the multifactorial nature of injury
occurrence (Kumar, 2001; Meeuwisse et al., 2007; Wiese-Bjornstal, 2009),
of which one of the most widely cited was developed by Williams and
Anderson (Fig \ref{fig:fig1}; Andersen and Williams, 1988; Williams and
Andersen, 1998).

Williams and Andersen's (Williams and Andersen, 1998) stress-injury
model proposed that when faced with a potentially stressful athletic
situation, an athlete's personality traits (e.g., hardiness, locus of
control and competitive trait anxiety), history of stressors (e.g.,
major life events and previous injuries) and coping resources (e.g.,
general coping behaviours) contribute to the injury response, either
interactively or in isolation. The stress response is central to the
model and reflects the bi-directional relationship between athletes'
appraisal of, and response to, a stressful athletic situation. The model
predicts that athletes who have a history of stressors, personality
traits that intensify the stress response and few coping resources, will
exhibit greater attentional (e.g., peripheral narrowing) and/or
physiological (e.g., increased muscle tension) responses that put these
individuals at greater injury risk.

Within Williams and Andersen's (Williams and Andersen, 1998) model,
major life events, which is a component of an athlete's history of
stressors, most consistently predicts injury occurrence (Williams and
Andersen, 2007); specifically, major life events with a negative, as
opposed to positive, valence (Passer and Seese, 1983; Maddison and
Prapavessis, 2005). Personality traits and coping resources have also
been found to predict injury with athletes more likely to sustain an
injury if they have poor social support and psychological coping skills,
and high trait anxiety and elevated competitive state anxiety; compared
to those with the opposing profile (Smith et al., 1990; Lavallée and
Flint, 1996; Ivarsson and Johnson, 2010).However, the amount of variance
explained by these psychosocial factors has been modest and typically
between 5 - 30\% (Galambos et al., 2005; Ivarsson and Johnson, 2010),
which indicates a likely interaction with other factors.

While the psychosocial factors in Williams and Andersen's (Williams and
Andersen, 1998) model have received the most research attention, less
insight into the mechanisms through which these factors are proposed to
exert their effect exists. To elaborate, the model suggests that
injuries are likely to occur through either increased physiological
arousal resulting in increased muscle tension and reduced flexibility or
attentional deficits caused by increased distractibility and peripheral
narrowing. However, to date, the research has largely focused on
attentional deficits (Andersen and Williams, 1999; Rogers and Landers,
2005; Wilkerson, 2012; Swanik et al., 2007). For example, Andersen and
Williams (Andersen and Williams, 1999) found athletes with high life
event stress coupled with low social support had greater peripheral
narrowing under stressful conditions compared to athletes with the
opposing profile; these athletes went on to sustain an increased number
of injuries during the following athletic season. Rodgers and Landers
(Rogers and Landers, 2005) further supported Andersen and Williams's
(Andersen and Williams, 1999) earlier findings by identifying that
peripheral narrowing under stress mediated 8.1\% of the relationship
between negative life events and injury.

Knowledge of the physiological factors (e.g., increased muscle tension
and reduced motor control) contributing to the remaining variance
between negative life events and athletic injury remains sparse
(cf.~Williams and Andersen, 1998). One challenge faced by researchers
addressing the sports injury problem through a psychological lens is the
multifactorial nature of injury, and the possible interaction with
physiological factors in the stress response (Meeuwisse et al., 2007;
Wiese-Bjornstal, 2009). For example, a large body of research has
suggested that training-related stress is also likely to contribute to
the stress response and injury occurrence (Lee et al., 2017; Djaoui et
al., 2017) and may account for the unexplained variance from the
psychological predictors. Appaneal and Perna (Appaneal and Perna, 2014)
proposed the biopsychosocial model of stress athletic injury and health
(BMSAIH) to serve as an extension to Williams and Andersen's (Williams
and Andersen, 1998) model. The BMSAIH enhances our understanding of the
mediating pathways between the stress response and injury alongside
other health outcomes and behavioral factors that impact sports
participation (Appaneal and Perna, 2014). The central tenet of the
BMSAIH is that psychosocial distress (e.g., negative life events) may
act synergistically with training-related stress as a result of
high-intensity and high-volume sports training, and ``widen the window
of susceptibility'' (Appaneal and Perna, 2014, 74) to a range of
undesirable health outcomes including illness and injury. Consequently,
the BMSAIH provides an important framework that has enhanced insight
into the multi-faceted nature of the injury process by building on
Williams and Andersen's (Williams and Andersen, 1998) model whilst
including other physiological markers of training-related stress.

Although research supporting the BMSAIH has mainly focused on the
relationship between hormonal responses to training and injury
occurrence (Perna and McDowell, 1995; Perna et al., 1997, 2003), other
training-related stress markers including heart rate variability
(Bellenger et al., 2016; Williams et al., 2017), muscle stiffness (Pruyn
et al., 2015) and postural stability (Romero-Franco et al., 2014) have
been reported to be associated with an increased risk of injury.
However, the largely mono-disciplinary approach to examining the role of
these markers has provided limited insight into the multifaceted
interactions with psychological factors that may contribute to injury
occurrence. Furthermore, a reliance on designs that capture a single
point of measurement precludes the assessment of intra as well as
inter-individual changes and the effect of the time interval between
measurement and injury occurrence on subsequent injuries (Johnson et
al., 2014). Such an approach fails to capture changes in both
psychological factors and physiological markers that may occur preceding
an injury. Importantly, a repeated measurement approach would enable an
assessing of how variables and their interactions change with respect to
time provide greater insight into the effect that repeated exposure to
major life events and other stress-related factors has on injury
occurrence.

Recently, Bittencourt et al.~(Bittencourt et al., 2016) advocated a move
away from studying isolated risk factors and instead, adopt a complex
systems approach in order to understand injury occurrence. Such an
approach posits that injury may arise from a complex ``web of
determinants'' (Bittencourt et al., 2016, 3), where different factors
interact in unpredictable and unplanned ways, but result in a global
outcome pattern of either adaptation or injury. Capturing the
uncertainty and complexity of the relationships between different
variables using an appropriate interdisciplinary analysis within the
framework of a complex systems approach is challenging. Bayesian network
(BN) modelling provides one solution by allowing the construction of
graphical probabilistic models using the underlying structure that
connects different variables (Scutari and Denis, 2014). The learned BN
structure can be used for inference by obtaining the posterior
probabilities of a particular variable for a given query (e.g., if the
value of variable A is x and the value of variable B is y, what is the
probability variable C of being value z?). Furthermore, unlike
regression or structural equation models, BN's do not distinguish
between dependent and independent variables when the underlying
relationship in the network may not be known (Olmedilla et al., 2018).
BN modelling subsequently provides a valuable but underused
interdisciplinary approach to investigating the complex and
unpredictable interactions of psychological and physiological factors
implicated in the injury process.

Using the frameworks provided by Williams and Andersen's (Williams and
Andersen, 1998) stress injury model and Appaneal's BMSAIH model, the aim
of this interdisciplinary study was to develop new understanding of the
multifaceted interactions of psychological and physiological
stress-related factors with injury occurrence. A prospective, repeated
measures design incorporating field-based physiological and
psychological measures combined with a BN modelling analysis was used to
address the study aim.

\section{Methods}

\subsection{Participants}

A total of 351 athletes (male: \emph{n} = 231, female: \emph{n} = 120)
were initially recruited for the study from a British University and
local sports clubs (Table \ref{tab:table1}). Participants had an average
age of 22.0 \(\pm\) 7.0 years and represented a range of team (football,
rugby, netball, cricket, lacrosse, basketball and field hockey) and
individual sports (athletics, tennis, weightlifting, gymnastics, judo,
swimming and golf. Participants self-rated competitive level ranged from
recreational to international standard. They were engaged in training
for their respective sports for at least five hours per week. A total of
126 (49.03\%) participants had sustained an injury in the 12 months
prior to the start of the study (male: \emph{n} = 114 {[}49\%{]},
female; \emph{n} = 48 {[}40\%{]}). All participants were injury free (no
modifications to their usual training routine due to a sport related
medical problem for a minimum of four weeks) at the start of the study.
Ethical approval was obtained from the University ethics committee prior
to the start of the study and all participants provided informed
consent.

\hypertarget{measures}{%
\subsection{Measures}\label{measures}}

\hypertarget{injury}{%
\subsubsection{Injury}\label{injury}}

Participants self-reported any injuries they sustained at each data
collection during the study period. An injury was defined as any sports
related medical problem causing the athlete to miss or modify their
usual training routine for at least 24 hours (Fuller et al., 2006, 2007;
Timpka et al., 2014). Minor scrapes and bruises that may require certain
modifications (e.g., strapping or taping) but did not limit continued
participation were not considered injuries (cf.~Appaneal et al., 2009).
Injury status (did / did not sustain an injury) served as the main
outcome measure.

\hypertarget{major-life-events}{%
\subsubsection{Major life events}\label{major-life-events}}

A modified version of the Life Events Survey for Collegiate Athletes
(LESCA) was used to measure participants' history of life event stress
(Petrie, 1992). The LESCA is the most widely used measure of major life
events for athletes in the sports injury literature. Modifications were
made to the LESCA to ensure the suitability of the items for the study
cohort (S1 Table). The LESCA comprises 69 items that reflect possible
life events that participants may have experienced. Example items
include, ``Major change in the frequency (increased or decreased) of
social activities due to participation in sport'', ``Major change in the
amount (more or less) of academic activity (homework, class time, etc)''
and ``Major change in level of athletic performance in actual
competition (better or worse)''. Participants were asked to rate the
perceived impact of each life event they had experienced within
12-months preceding the study onset on an 8-point Likert scale anchored
at -4 (\(extremely\;negative\)) and +4 (\(extremely\;positive\)).
Negative and positive life event scores were calculated by summing the
negative and positive scores, respectively. A score for total life
events was also calculated by summing the absolute values for both
negative and positive events. Petrie (1992) (Petrie, 1992) reported
test-retest reliabilities at 1-week and 8-weeks with values ranging from
.76 to .84 (p \textless{} .001) and .48 to .72 (p \textless{} .001)
respectively. Petrie also provided evidence of discriminant, convergent
and predictive validity. For this study, composite reliability (Fornell
and Larcker, 1981) was preferred to Cronbach's alpha as it does not
assume parallelity (i.e., all factor loadings are constrained to be
equal, and all error variances are constrained to be equal) and instead
takes into consideration the varying factor loadings of the items in the
questionnaire. The composite reliability for the LESCA in this study was
.84.

\hypertarget{reinforcement-sensitivity-theory-personality-questionnaire}{%
\subsubsection{Reinforcement Sensitivity Theory Personality
Questionnaire}\label{reinforcement-sensitivity-theory-personality-questionnaire}}

A revised version of the Reinforcement Sensitivity Theory Personality
Questionnaire (RST-PQ) was used to measure motivation, emotion,
personality and their relevance to psychopathology (Corr and Cooper,
2016). The revised version of the RST-PQ comprises 51 statements that
measure three major systems: Fight-Flight-Freeze System (FFFS; e.g., ``I
am the sort of person who easily freezes-up when scared''), Behavioural
Inhibition System (BIS; e.g., ``When trying to make a decision, I find
myself constantly chewing it over'') and four Behavioural Approach
System (BAS) factors; Reward Interest (e.g., ``I regularly try new
activities just to see if I enjoy them''), Goal Drive Persistence (e.g.,
``I am very persistent in achieving my goals''), Reward Reactivity
(e.g., ``I get a special thrill when I am praised for something I've
done well'') and Impulsivity (e.g., ``I find myself doing things on the
spur of the moment''). Participants rated each item on a scale from 1
(\(not\; at\; all\)) to 4 (\(highly\)) to reflect how well each
statement described their personality in general. The responses to items
associated with each subscale (FFFS, BIS, RI, GDP, RR and I) were summed
to give a total personality score that was subsequently used for further
analysis. The composite reliabilities for each subscale were; BIS =
0.92, FFFS = 0.77, GDP = 0.87, I = 0.71, RI = 0.77, RR = 0.81. Further
details regarding the revised RST are in S1 Appendix.

\hypertarget{heart-rate-variability}{%
\subsubsection{Heart rate variability}\label{heart-rate-variability}}

A Polar V800 heart rate monitor (HRM) and Polar H7 Bluetooth chest strap
(Polar OY, Finland) was used to collect inter-beat interval (IBI) data.
IBI recordings using the Polar V800 are highly comparable (ICC = 1.00)
with ECG recordings (Giles et al., 2016), which are considered the gold
standard for assessing HRV. In addition, HRV indices calculated from IBI
and ECG data have shown a strong correlation (r = .99) in athletes
(Caminal et al., 2018) and under spontaneous breathing conditions (Plews
et al., 2017).

\hypertarget{musculoskeletal-properties}{%
\subsubsection{Musculoskeletal
properties}\label{musculoskeletal-properties}}

A handheld myometer (MyotonPRO, Myoton AS, Tallinn, Estonia) was used to
measure passive muscle stiffness. The MyotonPRO is a non-invasive,
handheld device that applies a mechanical impulse of 0.40 N for 0.15 ms
perpendicular to the surface of the skin. The impulse causes natural
damped oscillations in the tissue, which are recorded by a three-axis
digital accelerometer sensor in the device. The raw oscillation signal
is then processed, and the stiffness parameter is calculated
(Agyapong-Badu et al., 2016). The MyotonPRO has previously been reported
to be a reliable and valid tool for the measurement of in-vivo tissue
stiffness properties (Chuang et al., 2013; Pruyn et al., 2016; Nair et
al., 2014), and has demonstrated good internal consistency (coefficient
of variation \textless{} 1.4\%) over sets of 10 repetitions (Aird et
al., 2012).

\hypertarget{postural-stability}{%
\subsubsection{Postural stability}\label{postural-stability}}

Postural stability was assessed with a modified version of the balance
error scoring system (mBESS) based on the protocol recommended by Hunt
et al. (2009). In total, each trial of the mBESS was performed without
shoes (McCrory et al., 2013) and included six stances in the following
order; dominant leg (DL; standing on the dominant foot with the
non-dominant foot at approximately 30-degrees of hip flexion and
45-degrees of knee flexion), non-dominant leg (NDL; standing on the
non-dominant foot with the dominant foot at approximately 30-degrees of
hip flexion and 45-degrees of knee flexion) and tandem leg stance (TS;
standing heel-to-toe with the non-dominant foot behind the dominant) on
firm and foam (Alcan airex AG, Sins, Switzerland) surfaces respectively
(Fig \ref{fig:fig2}). To determine leg dominance, participants were
asked their preferred leg to kick a ball to a target, and the chosen
limb was labelled as dominant (cf.~Cingel et al., 2017). Participants
were asked to maintain each stance for a total of 20-seconds.
Participants hands were placed on hips at the level of the iliac crests.
A Sony DSC-RX10 video camera (Sony Europe Limited, Surrey, United
Kingdom) was used to record each participants performance during the
mBESS.

The error identification criteria from the original BESS protocol was
used by the lead researcher who scored all the BESS trials. One error
was recorded if any of the following movements were observed during each
trial: a) lifting hands off iliac crests; b) opening eyes; c) stepping,
stumbling, or falling; d) moving the thigh into more than 30 degrees of
flexion or abduction; e) lifting the forefoot or heel; and f) remaining
out of the testing position for more than 5-seconds (Riemann et al.,
1999). A maximum score of 10 errors was possible for each stance.
Multiple errors occurring simultaneously were recorded as one error. A
participant was given the maximum score of 10 if they remained out of
the stance position for more than 5-seconds. A total score was
calculated by summing the total number of errors recorded on all stances
(DL, NLD and TS, on foam and firm surfaces). To assess the intra-rater
reliability, a single measurement, absolute agreement, two-way mixed
effects model for the intraclass correlation (ICC; Koo and Li, 2016) was
used on a sample of 40 participants from the first time point. The
test-retest scoring of BESS resulted in a ``good'' to ``excellent'' ICC
score (ICC = 0.93, 95\% confidence interval = 0.88 - 0.96), indicating
the scoring was reliable.

\hypertarget{procedure}{%
\subsection{Procedure}\label{procedure}}

At the start of the academic year (September), coaches of sports teams
at a British University and local sports clubs were contacted and
informed about the study. With the coaches' permission, the lead
researcher attended training sessions to inform athletes about the
overall purpose of the study and the requirements of participation.
Athletes who met the participation criteria and volunteered to take part
in the study were invited to attend scheduled testing sessions. A
repeated measures prospective cohort design was used to assess athletes'
major life events, stress-related physiological markers and injury
status over two consecutive 12-month periods. Each participant was asked
to attend a total of three data collections over a 12-month period, with
each data collection separated by a four-month interval (Fig
\ref{fig:fig3}). Participants provided informed consent before data
collection commenced.

For the first three data collections (T1, T2 and T3), participants
followed the same protocol in a specific order (Fig \ref{fig:fig4}). To
ensure all measures could be collected within a viable time-frame,
participants were separated into two groups. The first group completed
all computer-based measures followed by all physical measurements,
whereas the second group completed all physical measurements followed by
computer-based measures. Participants were randomly assigned to one of
the two groups and remained in those groups across all time points.

\hypertarget{questionnaires}{%
\subsubsection{Questionnaires}\label{questionnaires}}

The questionnaires, which included demographic information, the LESCA,
RST-PQ (T1, T2, T3) and injury status (T2, T3, T4) were completed
on-line (SurveyMonkey Inc., USA, www.surveymonkey.com). The instructions
for the LESCA were modified at T2 and T3 so that participants reported
major life events that had occurred since the previous testing session.
For injury reporting, participants were asked to record any injuries
that they had sustained since the last data collection. The data were
downloaded from surverymonkey.com and imported into R (R Core Team,
2019) for analysis.

\hypertarget{hrv}{%
\subsubsection{HRV}\label{hrv}}

To minimise potential distractions, participants were directed to a
designated quiet area in the laboratory where IBI data were recorded.
Participants were instructed to turn off their mobile devices to avoid
any interference with the Bluetooth sensor. Each chest strap was
dampened with water and adjusted so it fitted tightly but comfortably,
as outlined by Polar's guidelines. Participants were seated and asked to
remain as still as possible for the duration of the recording. No
attempt was made to control participants' respiratory frequency or tidal
volume (Denver et al., 2007). Inter-beat interval (IBI) data was
collected for 10-minutes at a sampling frequency of 1000 Hz.

Raw, unfiltered IBI recordings were exported from the Polar Flow web
service as a space delimited .txt file and imported into R (R Core Team,
2019) where the \emph{RHRV} package ({\textbf{???}}) was used to
calculate HRV indices. Raw IBI data was filtered using an adaptive
threshold filter, and the first 3-minutes and last 2-minutes of each
recording were discarded, leaving a 5-minutes window that was used to
calculate the root mean square of successive differences (RMSSD) in RR
intervals following the recommendations for short term IBI recordings
(Laborde et al., 2017; Malik et al., 1996). RMSSD was calculated as:

\begin{equation} 
  \overline{RR} = \frac{1}{N} \sum_{i=1}^{n} RR_i
\end{equation}

Where N is the length of the time series, and \(RR_i\) the RR interval
between beats \(i\) and \(i-1\), where each beat position corresponds to
the beat detection instant.

\hypertarget{muscle-stiffness}{%
\subsubsection{Muscle stiffness}\label{muscle-stiffness}}

To assess muscle stiffness, participants lay horizontally on a massage
bed and four testing sites were identified on each lower limb. The
muscle belly of the rectus femoris (RF), biceps femoris (BF), medial
gastrocnemius (MG) and lateral gastrocnemius (LG) sites were identified
using a visual-palpatory technique to determine the exact location of
each site (Chuang et al., 2012). The visual-palpatory technique required
the participant to contract the target muscle to aid the lead researcher
to visually identify the muscle. The participant was then asked to relax
the muscle and the muscle was palpated to locate the muscle belly. A
skin safe pen (Viscot all skin marker pen, Viscot Medical LLC, NJ) was
used to mark the testing site in the centre of the muscle belly.

After the eight testing sites had been identified, the testing end of
the MyotonPRO (diameter = 3 mm) was positioned perpendicular to the skin
on the testing site. A constant pre-load of 0.18 N was applied for
initial compression of subcutaneous tissues. The device was programmed
to deliver five consecutive impulses, separated by a one-second interval
(Morgan et al., 2018). For each impulse, the device computed passive
stiffness values, with the median of the five values being saved by the
device for further analysis. In accordance with Myoton.com, a set of
five measurements with a coefficient of variation (CV) of less than 3\%
was accepted. Sets of measurements above 3\% were measured again to
ensure data reliability. Measurements were uploaded using MyotonPRO
software and imported in R (R Core Team, 2019) for further analysis. For
each participant, the sum of all eight testing sites was calculated to
provide a total lower extremity stiffness score and was used for further
analysis

\hypertarget{postural-stability-1}{%
\subsubsection{Postural stability}\label{postural-stability-1}}

Instructions for the mBESS were read to each participant and a
demonstration of the positions was provided by the research assistant.
For each position, participants were instructed to close their eyes,
rest their hands on their iliac crests and remain as still as possible
for 20-seconds. Participants were instructed to return to the testing
position as quickly as possible if they lost their balance. The video
recording was started prior to the first stance position and stopped
after all stances had been completed. Each completed mBESS protocol took
approximately four-minutes. Only one trial was performed to avoid
familiarisation effects across the repeated measurement (cf.~Valovich et
al., 2003). The video recordings for each participant were imported from
the recording equipment (Sony DSC-RX10) and the lead researcher scored
each trial using the error identification criteria.

\hypertarget{data-analysis}{%
\subsection{Data Analysis}\label{data-analysis}}

Two Bayesian Networks (BN) were used to explore the relationships
between the psychological measures, physiological markers of stress and
sports injury. A BN is a graphical representation of a joint probability
distribution among a set of random variables, and provides a statistical
model describing the dependencies and conditional independences from
empirical data in a visually appealing way (Scutari and Denis, 2014). A
BN consists of arcs and nodes that together are formally known as a
directed acyclic graph (DAG), where a node is termed a parent of a child
if there is an arc directed from the former to the latter (Fig
\ref{fig:fig5}; Pearl, 1988). However, the direction of the arc does not
necessarily imply causation, and the relationship between variables are
often described as probabilistic instead of casual (Scutari and Denis,
2014). The information within a node can be either continuous or
discrete, and a complete network can contain both continuous and
discrete nodes; however, discrete networks are the most commonly used
form of BN (Chen and Pollino, 2012). In discrete networks, conditional
probabilities for each child node are allocated for each combination of
the possible states in their parent nodes and can be used to assess the
strength of a dependency in the network.

In order to use discrete networks, continuous variables must first be
split into categorical levels. When there are a large number of
variables in the network, limiting the number of levels has the benefit
of producing a network that is more parsimonious in terms of parameters.
For example, a network with 10 variables each with two levels has 100
(10\^{}2) possible parameter combinations, however the same network with
three levels has 1000 (10\^{}3) possible parameter combinations, the
latter being significantly more computationally expensive. Using a
larger number of splits in the data also comes at a cost of reducing the
statistical power in detecting probabilistic associations, and reduces
the precision of parameter estimates for the probabilistic associations
that are detected because it reduces the sample-size-to-parameters ratio
(Scutari and Denis, 2014). Typically, no more than three levels have
been used in Bayesian networks in the sports injury literature
(Olmedilla et al., 2018; Ruiz-Pérez et al., 2019)

Learning the structure of the network is an important step in BN
modelling. The structure of a network can be constructed using expert
knowledge and/or data-driven algorithm techniques (e.g., search and
score, such as hill climbing and gradient descent algorithms; Scutari \&
Denis, 2014). The learned structure can then be used for inference by
querying the network\footnote{The term ``query'' in relation to Bayesian
  Networks stems from Pearl's expert systems theory (1988). A query can
  be submitted to an expert (in this case, the network is the expert) to
  get an opinion, the expert then updates the querier's beliefs
  accordingly. Widely used texts on Bayesian Network analysis (Koller
  and Friedman, 2009) have widely adopted the terminology in favour of
  that used in traditional statistics.} and obtaining the posterior
probabilities of a particular node for a given query. The posterior
distribution can be obtained by \(Pr(X|E,B) = Pr(X|E,G,\Theta)\), where
the learned network \(B\) with structure \(G\) and parameters
\(\Theta\), are investigated with new evidence \(E\) using the
information in \(B\) (Scutari and Denis, 2014). In the example network
presented in Fig \ref{fig:fig5}, new values assigned to each of the
parent nodes (e.g., both set to ``Low'') could be used to investigate
what effect the new information has on the state of the child node
(conditional probability of a particular state of the child node). In a
more complex network containing many nodes, the outcome of a particular
node can be assessed conditional on the states of any subset of nodes in
the network. BNs therefore provide a unique and versatile approach to
modelling a set of variables to uncover dependency structures within the
data.

BNs have recently been used in the sport psychology literature
(Olmedilla et al., 2018; Fuster-Parra et al., 2017) and offer several
benefits over traditional statistical analysis. For example, predictions
can be made about any variable in the network, rather than there being a
distinction between dependent and independent variables in the data,
such as in linear regression models that are often used within the sport
psychology literature (Olmedilla et al., 2018; Bittencourt et al.,
2016). Furthermore, the structure of a network can be obtained from both
empirical data \emph{and} prior knowledge about the area of study; the
latter being particularly useful when there are a large number of
variables in the network, or only a small number of observations are
available in the data (Xiao-xuan et al., 2008). In such instances, a
purely data driven approach to learning the network would be
time-consuming due to the large parameter space, and inefficiency at
identifying an approximation of the true network structure. Prior
knowledge about dependencies between variables can therefore be included
in the network structure, while still allowing a data driven approach
for unknown dependencies, to improve the overall computation of the
network structure (Heckerman et al., 1995; Xu et al., 2015). The
following sections detail the steps taken in the current study to
firstly prepare the data for each network, and then obtain the structure
of each network that was used for further inference.

\hypertarget{first-network}{%
\subsubsection{First network}\label{first-network}}

\hypertarget{data-preparation}{%
\paragraph{Data Preparation}\label{data-preparation}}
